Icicle: write once, run once

### 2016-06-09 2016fhpc/figures/Bench-All.tex
minor changes

\begin{tikzpicture}
\begin{axis}[
	ylabel=MB/s,
  ymin=0, ymax=1000,
	enlargelimits=0.01,
	ybar interval=0.7,
  symbolic x coords={R, wc, 1 CPU, grep, cat, 32 CPU, end},
  width=8.5cm, height=5cm,
]
\addplot coordinates {(R,6.0) (wc,182.6) (1 CPU,186.6) (grep,261) (cat,878) (32 CPU,908) (end,0) };
% \addplot coordinates {(R,6.0) (wc,182.6) (CPU1,186.6) (grep,261) (wc -l,852) (cat,878) (Run 372,908) (Run 1,1029) (end,0) };

>>>>

\begin{tikzpicture}
\begin{axis}[
	ylabel=MB/s,
  ymin=0, ymax=1000,
	enlargelimits=0.01,
	ybar interval=0.7,
  symbolic x coords={R, wc, 1 CPU, grep, cat, 32 CPU, end},
  width=8.5cm, height=6.5cm,
]
\addplot coordinates {(R,6.0) (wc,182.6) (1 CPU,186.6) (grep,261) (cat,878) (32 CPU,908) (end,0) };
% \addplot coordinates {(R,6.0) (wc,182.6) (CPU1,186.6) (grep,261) (wc -l,852) (cat,878) (Run 372,908) (Run 1,1029) (end,0) };

### 2016-06-09 2016fhpc/section/01-Introduction.tex
minor changes

Such a joint query can be converted to a back-end language like SQL, but doing so by hand is tedious and error prone. As the three queries use different filter predicates we cannot use a single @SELECT@ statement and a @WHERE@ expression to implement the filter. We must instead lift each predicate to an expression-level conditional:
\begin{code}
  SELECT COUNT(IF(open > close, 1, 0))
       , COUNT(IF(open < close, 1, 0))
       , SUM  (IF(open > close, open, 0))
       / COUNT(IF(open > close, 1, 0))
  FROM stocks;
\end{code}
As we see, the result of query fusion tends to have many common sub expressions, and we wish to guarantee that the duplicates in the fused result are eliminated.

>>>>

Such a joint query can be converted to a back-end language like SQL, but doing so by hand is tedious and error prone. As the three queries use different filter predicates we cannot use a single @SELECT@ statement and a @WHERE@ expression to implement the filter. We must instead lift each predicate to an expression-level conditional and compute the count by summing the conditional:
\begin{code}
  SELECT SUM(IF(open > close, 1,    0))
       , SUM(IF(open < close, 1,    0))
       , SUM(IF(open > close, open, 0))
       / SUM(IF(open > close, 1,    0))
  FROM stocks;
\end{code}
As we see, the result of query fusion tends to have many common sub expressions, and we wish to guarantee that the duplicates in the fused result are eliminated.

### 2016-06-09 2016fhpc/section/02-Source.tex
Copy edits to intermediate language
minor changes

\eject
% ---------------------------------------------------------
\section{Elements and Aggregates}
To allow incremental computation all Icicle queries must execute in a single pass over the input stream.
Sadly, not all queries \emph{can} be executed in a single pass: the key examples are queries that require random access indexing, or otherwise need to access data in an order different to what the stream provides.
However, as we saw in the introduction, although a particular \emph{algorithm} may be impossible to evaluate in a streaming fashion, the desired \emph{value} may well be computable, if only we had a different algorithm.
Here is the unstreamable example from the introduction again:
\begin{code}
  table kvs { key : Date; value : Real }
  query avg = let k = last key
              in  filter (key == k) of mean value;
\end{code}

@@

% ---------------------------------------------------------
\subsection{Finite Streams and Synchronous Data Flow}
In contrast to synchronous data flow languages such as {\sc Lustre}~\cite{halbwachs1991synchronous}, all streams processed by Icicle are conceptually finite in length. Icicle is fundamentally a query language, which queries finite tables of data held in a non-volatile store, but does so in a streaming manner. Lustre operates on conceptually infinite streams, such as those found in real-time control systems (like to fly airplanes). In Icicle, the ``last'' element in a stream is the last one that appears in the table on disk. In Lustre, the ``last'' element in a stream is the one that was most recently received. If the unstreamable query from \S2 was converted to Lustre syntax then it would execute, but the filter predicate would compare the last key with the most recent key from the stream, which is the key itself. The filter predicate would always be true, and the query would return the mean of the entire stream. Applying the Icicle type system to our queries imposes the natural stage restriction associated with finite streams, so there are distinct ``during'' (element) and ``after'' (aggregate) stages.

@@

% ---------------------------------------------------------
\subsection{Type System}
The typing rules for Icicle are given in Figure~\ref{fig:source:type:exp}.
The judgment form $\Typecheck{\Gamma}{\mi{e}}{\TauMode}$ associates an expression $\mi{e}$ with its type $M$ under context $\Gamma$.
The judgment form $\TypecheckP{\mi{p}}{\TauFun}$ associates a primitive with its function type.
The judgment form $\TypecheckApp{\TauFun}{\ov{\TauMode}}{\TauMode}$ is used to lift function application to modal types: a function type applied to a list of modal argument types produces a result type and matching mode.
The judgment form $\TypecheckS{\Gamma}{\mi{Def}}{\Gamma}$ takes an input environment and function or query, and produces an environment containing the function or query name and its type.
Finally, $\TypecheckS{}{\mi{Top}}{\Gamma}$ takes a top-level definition with a table, functions and queries, and produces a context containing only the types of the queries.

@@

Rules TcPrimApp and TcFunApp produce the type a primitive or function applied to its arguments. Rule TcLet is standard.

@@

Rules EBoxStream and EBoxFold lift constant values to stream results and aggregate results respectively. To lift a constant to a stream result we produce a meta-function that always returns the value. To lift a constant to an aggregate result we set the update meta-function to return a dummy, and have the eject meta-function return the value of interest.

>>>>

\eject
% ---------------------------------------------------------
\section{Elements and Aggregates}
\label{s:ElementsAndAggregates}
To allow incremental computation all Icicle queries must execute in a single pass over the input stream.
Sadly, not all queries \emph{can} be executed in a single pass: the key examples are queries that require random access indexing, or otherwise need to access data in an order different to what the stream provides.
However, as we saw in the introduction, although a particular \emph{algorithm} may be impossible to evaluate in a streaming fashion, the desired \emph{value} may well be computable, if only we had a different algorithm.
Here is the unstreamable example from the introduction again:
\begin{code}
  table kvs { key : Date; value : Real }
  query avg = let k = last key
              in  filter (key == k) of mean value;
\end{code}

@@

% ---------------------------------------------------------
\subsection{Finite Streams and Synchronous Data Flow}
In contrast to synchronous data flow languages such as {\sc Lustre}~\cite{halbwachs1991synchronous}, the streams processed by Icicle are conceptually finite in length. Icicle is fundamentally a query language, which queries finite tables of data held in a non-volatile store, but does so in a streaming manner. Lustre operates on conceptually infinite streams, such as those found in real-time control systems (like to fly airplanes). In Icicle, the ``last'' element in a stream is the last one that appears in the table on disk. In Lustre, the ``last'' element in a stream is the one that was most recently received. If the unstreamable query from \S2 was converted to Lustre syntax then it would execute, but the filter predicate would compare the last key with the most recent key from the stream, which is the key itself. The filter predicate would always be true, and the query would return the mean of the entire stream. Applying the Icicle type system to our queries imposes the natural stage restriction associated with finite streams, so there are distinct ``during'' (element) and ``after'' (aggregate) stages.

@@

% ---------------------------------------------------------
\subsection{Type System}
The typing rules for Icicle are given in Figure~\ref{fig:source:type:exp}.
The judgment form $\Typecheck{\Gamma}{\mi{e}}{\TauMode}$ associates an expression $\mi{e}$ with its type $M$ under context $\Gamma$.
The judgment form $\TypecheckP{\mi{p}}{\TauFun}$ associates a primitive with its function type.
The judgment form $\TypecheckApp{\TauFun}{\ov{\TauMode}}{\TauMode}$ is used to lift function application to modal types: a function type applied to a list of modal argument types produces a result type and matching mode.
The judgment form $\TypecheckS{\Gamma}{\mi{Def}}{\Gamma}$ takes an input environment and function or query, and produces an environment containing the function or query name and its type.
Finally, $\TypecheckS{}{\mi{Top}}{\Gamma}$ takes a top-level definition with a table, functions and queries, and produces a context containing the types of all the definitions.

@@

Rules TcPrimApp and TcFunApp produce the type of a primitive or function applied to its arguments. Rule TcLet is standard.

@@

Rules EBoxStream and EBoxFold lift constant values to stream results and aggregate results respectively. To lift a constant to a stream result we produce a meta-function that always returns the value. To lift a constant to an aggregate result we set the update meta-function to return a dummy value, and have the eject meta-function return the value of interest.

### 2016-06-09 2016fhpc/section/03-Core.tex
minor changes

Inlining the above functions into the query from \label{s:ElementsAndAggregates} yields the following:
% This is guaranteed to terminate because no recursion is allowed in function definitions.
\begin{code}
  query avg
   =    let lst = (fold l = NO_DATE then key)
     in let map = group key of
                  ( (fold s = 0 then s + value)
                  / (fold c = 0 then c + 1) )
     in let ret = lookup lst map
     in     ret
\end{code}

@@

For the @map@ binding, each fold accumulator inside the body of the @group@ construct is associate with its own finite map. The @s@ accumulator is associated with map @gS@, and the @c@ accumulator with @gC@. Each time we receive a row from the table the accumulator associated with the @key@ is updated, using the default value @0@ if an entry for that key is not yet present. After we have processed the entire table we divide each sum with its corresponding count to yield a map of means for each key.
\begin{alltt}
  folds  { gS  : Map Date Real = mapEmpty
           then mapUpdate gS key 0 (\(\lambda\)s. s + value)

         ; gC  : Map Date Real = mapEmpty
           then mapUpdate gC key 0 (\(\lambda\)c. c + 1) }

  after  { map : Map Date Real
            = mapMap (\(\lambda\)s c. s / c) (mapZip gS gC) }
\end{alltt}

@@

To combine the plans from each binding we simply concatenate the corresponding parts of the anatomy. To fuse multiple plans we freshen the names of each binding and also concatenate the corresponding parts of the anatomy. The single-pass restriction on queries makes the fusion process so simple. The single-pass restriction implies that there are no fusion-preventing dependencies between any two query plans.

Given a fused query plan we the convert it to an imperative loop nest in a similar way to our prior work on flow fusion~\cite{lippmeier2013data}.

>>>>

Inlining the above functions into the query from \S\ref{s:ElementsAndAggregates} yields the following:
% This is guaranteed to terminate because no recursion is allowed in function definitions.
\begin{code}
  query avg
   =    let lst = (fold l = NO_DATE then key)
     in let map = group key of
                  ( (fold s = 0 then s + value)
                  / (fold c = 0 then c + 1) )
     in let ret = lookup lst map
     in     ret
\end{code}

@@

For the @map@ binding, each fold accumulator inside the body of the @group@ construct is associated with its own finite map. The @s@ accumulator is associated with map @gS@, and the @c@ accumulator with @gC@. Each time we receive a row from the table the accumulator associated with the @key@ is updated, using the default value @0@ if an entry for that key is not yet present. After we have processed the entire table we divide each sum with its corresponding count to yield a map of means for each key.
\begin{alltt}
  folds  \{ gS  : Map Date Real = mapEmpty
           then mapUpdate gS key 0 (\(\lambda\)s. s + value)

         ; gC  : Map Date Real = mapEmpty
           then mapUpdate gC key 0 (\(\lambda\)c. c + 1) \}

  after  \{ map : Map Date Real
            = mapMap (\(\lambda\)s c. s / c) (mapZip gS gC) \}
\end{alltt}

@@

To combine the plans from each binding we simply concatenate the corresponding parts of the anatomy. To fuse multiple plans we freshen the names of each binding and also concatenate the corresponding parts of the anatomy. The single-pass restriction on queries makes the fusion process so simple, because it ensures that there are no fusion-preventing dependencies between any two query plans.

Given a fused query plan we then convert it to an imperative loop nest in a similar way to our prior work on flow fusion~\cite{lippmeier2013data}.

### 2016-06-09 2016fhpc/section/04-Benchmarks.tex
minor changes

We ran all the Unix utilities with unicode decoding disabled using @LANG=C LC_COLLATE@ for maximum performance. The input data does not contain unicode characters. 
We used an Amazon EC2 @c3.8xlarge@ with 32 CPUs, 60GB of RAM, and striped, RAIDed SSD storage. The fused Icicle version significantly outperformed the R version of the queries, and the single-threaded version was on par with @wc@, while only a little slower than @grep@. This is despite the fact that the Icicle queries perform more computational work than than @wc@ and @grep@. By using multiple processors, we were able to scale up to perform as well as @cat@, approaching the disk speed.
The R code is single threaded and would require at least 150 processors to reach similar speeds, assuming perfect scaling.
These results give us confidence that our distributed implementation will be fast as well as scalable~\cite{mcsherry2015scalability}.

@@

Figure~\ref{fig:bench:queries} shows how the total througput scales as the number of fused queries is increased. For each number of queries, we ran two versions of the fused result: one version that wrote the output to disk, and the other that piped the result to @/dev/null@. The graph shows the throughput of the disk version decreasing roughly linearly in the number of queries, while the version ignoring the output remains constant. This suggests that we are IO bound on the write side. The time spent evaluating the queries themselves is small relative to our current IO load, and we are curently scaling up our system to use a larger query set.

>>>>

We ran all the Unix utilities with unicode decoding disabled using @LANG=C LC_COLLATE@ for maximum performance. The input data does not contain unicode characters. 
We used an Amazon EC2 @c3.8xlarge@ with 32 CPUs, 60GB of RAM, and striped, RAIDed SSD storage. The fused Icicle version significantly outperformed the R version of the queries, and the single-threaded version was on par with @wc@, while only a little slower than @grep@. This is despite the fact that the Icicle queries perform more computational work than than @wc@ and @grep@. By using multiple processors, we were able to scale up to perform as well as @cat@, approaching the disk speed.
The memory usage of Icicle starts at around 200MB of RAM for a single thread, but as more threads are added approaches 15MB per thread.
The memory usage is constant in the input size and depends on the number of queries.
The R code is single threaded and would require at least 150 processors to reach similar speeds, assuming perfect scaling.
These results give us confidence that our distributed implementation will be fast as well as scalable~\cite{mcsherry2015scalability}.

@@

Figure~\ref{fig:bench:queries} shows how the total read througput scales as the number of fused queries is increased. For each number of queries, we ran two versions of the fused result: one version that wrote the output to disk, and the other that piped the result to @/dev/null@. The graph shows the throughput of the disk version decreasing roughly linearly in the number of queries, while the version ignoring the output remains constant. This suggests that we are IO bound on the write side. The time spent evaluating the queries themselves is small relative to our current IO load, and we are curently scaling up our system to use a larger query set.
